---
title: "DATA MAIPULATION"
author: "Uyen Vu (uyen.vu@phdstudent.hhs.se)"
date: "6/13/2022"
output: 
  word_document:
    reference_docx: "assets/sse-word-template.docx"
    toc: true
    toc_depth: 2
monofont: 'Consolas' #How to find other fonts?
---

\newpage

# PACKAGES in R
## Install packages
- Sources of packages:
  + CRAN repository: Packages here are at least reviewed by someone before being published
  + install directly from **github**: In github, we can find the latest packages or sometimes even those that have not been published in the official repository, these packages are **NOT reviewed by anyone**. Hence, there is no guarantee that the package works.

- Syntax: *install.packages("tidyverse")*
  + *tidyverse* is the name of the package.
  + replace *tidyverse* by any other name.
  
- From **github**
  + *remotes::install_github("user/repository")*
  + *user* is the one who posted the package
  + *repository* is the name of the package

## Use packages in R
- Syntax: *package_name::function_name()*. This is the best way to call.
- using *library()*. This is used when you use a lot of different functions from the same package in your script. You load the library for once, and then you do not need to call the name of the package whenever you use the functions from the packages.

# IMPORTING, EXPORTING, SELECTING
## Importing
- Before, there are different ways to deal with importing, depending on which type of file you want to import
- **rio** package helps to reduce the workload. It allows more than 30 different datafile format, e.g. csv, excel, SAS, SPSS, Stata, matlab, json, etc.
- syntax: *rio::import("")*
- We can use the example with 2 datafiles as below. This is PISA survey (2018) from OECD website.
  + cy07_msu_sch_qqq.sas7bdat: SAS format
  + CY07_MSU_SCH_QQQ.sav: SPSS file

```{r}
# Dataset in SAS format
pisa_sas <- rio::import("data/cy07_msu_sch_qqq.sas7bdat")
```

```{r}
# Dataset in SPSS format
pisa_spss <- rio::import("data/CY07_MSU_SCH_QQQ.sav")
```

## Exporting
- We can use base function or **rio** package
- Base function:
  + syntax: *saveRDS(object,"path/file.Rds")*
  + this is standard format to save data in R. This avoids loss of information.
- However, if you want to export data to some other format (e.g. excel), use *rio* package:
  + syntax: *rio::export(object,"path/file.xlsx")*
  + we can replace *xlsx* by other formats
- Example (with data used in **importing** section):

```{r}
saveRDS(pisa_sas, "data/pisa_sas.Rds")
```

```{r}
#saving to excel file
rio::export(pisa_sas, "data/pisa_sas.xlsx")
```

```{r}
#saving for STATA
rio::export(pisa_sas, "data/pisa_sas.dta")
```

## Selecting
- For now, you can just copy the code below and run (We are simulating a dataset, which we learn later in details)

```{r}
set.seed(123)
df <- data.frame(id = c(1:20),
                     name = stri_rand_strings(n = 20, length = 7, pattern = "[A-Za-z0-9]"),
                     class = sample(x=c("A", "B", "C"), size = 20, replace = TRUE, prob = c(0.3,0.3,0.4)),
                     gender = sample(x = c("female","male"), size = 20, replace = TRUE, prob = c(0.5,0.5)),
                     district = sample(x = c(3:8), size = 20, replace = TRUE, prob = c(0.2,0.1,0.25,0.15,0.2,0.1)),
                     campus = sample(x = c("Yes","No"), size = 20, replace = TRUE, prob = c(0.3,0.7)),
                     transport = sample(x = c("shuttle bus", "bicycle", "car"), size = 20, replace = TRUE, prob = c(0.55,0.2,0.25)),
                     math = round(rnorm(n = 20, mean = 7.2, sd = 0.6), 1),
                     english = round(rnorm(n = 20, mean = 8.1, sd = 0.5), 1)
)
View(df)
```

- We are going to get 4 columns: *id, name, class, math*. There are 2 ways:
  + Base function in R:
  
```{r}
df1 <- df[, c("id","name", "class", "math")]
#or:
df1 <- df[, c(1:3,8)]
#or:
df1 <- df[, names(df) %in% c("id","name", "class", "math")]
```

  + use package **dplyr**

```{r}
df2 <- dplyr::select(df, id, name, class, math)
```

#EXAMINING DATA
## Data structure and data type
- 5 data structures: atomic vector (elements of the same type), list, matrix, dataframe, factors
- 6 data types: integer, numeric, character, logical (boolean), complex, raw
- We continue with the simulated data in *selecting* section


```{r}
#Overally check data structure of dataframe
str(df)
```

- To check data type, we can use either *class()* or *typeof()*
  + *class()*: high level, the type of data as it appears to you
  + *typeof()*: low level, how R is storing the data
- Reference for data type: https://mgimond.github.io/ES218/Week02a.html
```{r}
#Check data type
class(df$math) #numeric
typeof(df$math) #double
is.numeric(df$math) #TRUE
#Check data type of all variables:
sapply(df, class)
```

## The use of length() function
- return the length of an R object (vector, list, string)
- Length of a vector:

```{r}
x <- c(8, 17, 23, 93, - 20, 15, 13, 55, 29, - 84)
length(x) #10 - this counts all elements in object x
```

- Length of a list

```{r}
set.seed(1357)                                     # Set seed for reproducibility
my_list <- list()                                  # Create empty list
my_list[[1]] <- c(5, 7, 1)                         # List element 1
my_list[[2]] <- rnorm(8, 5, 10)                    # List element 2
my_list[[3]] <- data.frame(x1 = 1:3, x2 = 4:6)     # List element 3
my_list
length(my_list) #3 - this counts the number of entries of our list, not the length of each element of the list
length(my_list[[2]]) #8 - this counts the number of elements of list[[2]]
```

- Length of a string

```{r}
x_string <- "Hello this is a string"
length(x_string) #1 - this means 1 string, it does not count the number of characters in the string
length(unlist(strsplit(x_string, ""))) #22
nchar(x_string) #22
```

- Length function applied in a data frame (use the simulated data in **selecting** section)

```{r}
length(df) #return number of column in the df
length(df$name) #return number of row (obs) in the df
length(unique())
dplyr::n_distinct()
```

## Data frame
- Create data frame using *data.frame()*
- some useful functions to deal with data frame:
  + *head()* & *tail()*
  + *dim()*
  + *nrow()* & *ncol()*
  + *str()*
  + *names()* or *colnames()*

## A note on tibble
- *tibble* is an alternative to the dataframe that has been created as part of the tidyverse package
- In most cases, there is no difference between *tibble* and *dataframe*
- *tibble* differs from *dataframe* when:
  + displaying data: while *dataframe* allows printing as many as possible (decided by *max.print* option in R environment), *tibble* by default prints the first 10 rows as as many columns as will fit in the window (readable)
  + matching: with *dataframe* when using $ to reference columns, partial names will be matched if the reference is not exact. This might be a problem when we have a typo mistake
- Create a tibble using syntax: *tibble::as.tibble(x)*

# DATA MANIPULATION
We use the simulated data:

```{r}
set.seed(123)
df <- data.frame(id = c(1:20),
                 name = stringi::stri_rand_strings(n = 20, length = 7, pattern = "[A-Za-z0-9]"),
                 class = sample(x=c("A", "B", "C"), size = 20, replace = TRUE, prob = c(0.3,0.3,0.4)),
                 gender = sample(x = c("female","male"), size = 20, replace = TRUE, prob = c(0.5,0.5)),
                 district = sample(x = c(3:8), size = 20, replace = TRUE, prob = c(0.2,0.1,0.25,0.15,0.2,0.1)),
                 campus = sample(x = c("Yes","No"), size = 20, replace = TRUE, prob = c(0.3,0.7)),
                 transport = sample(x = c("shuttle bus", "bicycle", "car"), size = 20, replace = TRUE, prob = c(0.55,0.2,0.25)),
                 math = round(rnorm(n = 20, mean = 7.2, sd = 0.6), 1),
                 english = round(rnorm(n = 20, mean = 8.1, sd = 0.5), 1)
)
```

## Extract columns from a dataframe
(see the **selecting* section above)

## Find the order of the column in the dataframe
<ADD>
## Remove dataframe and variables
- If we have created the dataframe that contains the variables of our interest, we can remove the prior dataset to save memory
- Completely remove a dataset from the environment: *rm(x)*

```{r}
rm(df1,df2)
```

- Remove some specific columns
  + using base function:

```{r}
#select 1 column
df1 <- df[, "name"]
#select multiple column
df1 <- df[, names(df) %in% c("transport","district")]
#remove multiple column
df[, names(df) %in% c("transport","district")] <- NULL
```

  + using **tidyverse** package:

```{r}
df1 <- dplyr::select(df, -transport, -district)
```

## Rename column:
- Using base R function:
```{r}
names(df)[2] <- "full_name"
names(df)[names(df)=="name"] <- "full_name"
#Rename all columns:
names(df1) <- c("id","full_name","class","gender","district","campus","transport","math","english")
```

- Using **tidyverse** package:

```{r}
df1 <-dplyr::rename(df1, full_name = name)
#multiple columns:
df1 <- dplyr::rename(df1,
                     full_name = names,
                     sex = gender)
```

- Making uppercase for all columns:

```{r}
df1 <- dplyr::rename_with(df1,toupper)
```

## Value labels:

```{r}
#Check factor levels:
levels(df$gender) # "female male" meaning female is assigned as 1 and male is assigned as 2 (alphabetical order)
nlevels(df$gender)

#Change factor levels:
df$gender <- factor(x=df$gender,levels=c("male","female")) # now male is assigned as 1
```

## Extract a random sample from a dataframe

```{r}
df <- df[sample(nrow(df), 5000), ] #5000 random rows you want to extract
```

## Gather data

```{r}
#Simulate data
set.seed(123)
df <- data.frame(id = c(1:10),
                 name = stringi::stri_rand_strings(n = 10, length = 7, pattern = "[A-Za-z0-9]"),
                 y1999 = round(rnorm(n=10, mean = 11, sd = 4), 1),
                 y2000 = round(rnorm(n=10, mean = 12.4, sd = 3.3), 1)
                 )
#Gather data
df <- tidyr::gather(df, key = "year", value = "rating", y1999:y2000)
```

## Spreading data

```{r}
df <- tidyr::spread(df, key = "year", value = "wage")
```

## Cleaning data
- This includes *filter()*, *mutate()*, *summarize()*, *arrange()*, *group_by()*
- We use below data simulation:

```{r}
set.seed(123)
df <- data.frame(id = c(1:20),
                 name = stringi::stri_rand_strings(n = 20, length = 7, pattern = "[A-Za-z0-9]"),
                 class = sample(x=c("A", "B", "C"), size = 20, replace = TRUE, prob = c(0.3,0.3,0.4)),
                 gender = sample(x = c("female","male"), size = 20, replace = TRUE, prob = c(0.5,0.5)),
                 district = sample(x = c(3:8), size = 20, replace = TRUE, prob = c(0.2,0.1,0.25,0.15,0.2,0.1)),
                 campus = sample(x = c("Yes","No"), size = 20, replace = TRUE, prob = c(0.3,0.7)),
                 transport = sample(x = c("shuttle bus", "bicycle", "car"), size = 20, replace = TRUE, prob = c(0.55,0.2,0.25)),
                 math = round(rnorm(n = 20, mean = 7.2, sd = 0.6), 1),
                 english = round(rnorm(n = 20, mean = 8.1, sd = 0.5), 1)
)
```

### filter()
- Some useful operators: **&, |, ==, !=, <, >, <=, >=**
```{r}
#Base R
df1 <- df[df$district == 3, ]
df1 <- df[(df$district == 3) | (df$district == 8), ]
#tidyverse
df1 <- dplyr::filter(df, district == 3 | district == 8)
```
- If you filter with & condition using **tidyverse**, you don't need to use the operator **&**, just separate the conditions by comma

### mutate()
- This is to create a new variable in a dataset

```{r}
df1 <- dplyr::mutate(df, average = ((math + english) / 2))
```

### arrange()
- Arrange the rows

```{r}
#Base R: decendent by math and ascendent by english
df1 <- df[order(df[, "english"], -df[, "math"]), ]
#tidyverse
df2 <- dplyr::arrange(df, english, -math)
```

- Arrange the columns

```{r}
#We use select() with everything() argument
df3 <- dplyr::select(df, english, math, everything())
```

### group_by()

```{r}
#Base R: 
aggregate(math ~ class, df, mean)
mean <- tapply(df$math, df$gender, mean)
sd <- tapply(df$math, df$gender, sd)
cbind(mean,sd)
#tidyverse:
df %>% group_by(class) %>% summarize(mean = mean(math))
```

### Recode variable

```{r}
#Base R:

#tidyverse
df$gender <- dplyr::recode(df$gender, "female" = 1, "male" = 0)
```

### summarize()
- Some popular summary statistics functions: *mean()*, *median()*, *sd()*, *cor()*, *quantile()*, *IQR()*, *table()*
- If you want to exclude the missing value, use **na.rm = TRUE**

```{r}
table(df$district)
tmp <-  as.data.frame(table(df$district))
rio::export(tmp, "data/district.xlsx")
```

## Missing values
- Identify missing values:

```{r}
#see what cases are missing
df_missing <- is.na(df$class)
#count numbers of cases that are missing
table(df_missing)
```

- Removing missing values:

```{r}
df <- df[complete.cases(df), ]
# OR
df <- df[!is.na(df$class), ]
```

- Recode missing values:

```{r}
#Base R
#df[df$gender == ]
```

# BASIC STATISTICS

```{r}
set.seed(123)
df <- data.frame(id = c(1:20),
                 name = stringi::stri_rand_strings(n = 20, length = 7, pattern = "[A-Za-z0-9]"),
                 class = sample(x=c("A", "B", "C"), size = 20, replace = TRUE, prob = c(0.3,0.3,0.4)),
                 gender = sample(x = c("female","male"), size = 20, replace = TRUE, prob = c(0.5,0.5)),
                 district = sample(x = c(3:8), size = 20, replace = TRUE, prob = c(0.2,0.1,0.25,0.15,0.2,0.1)),
                 campus = sample(x = c("Yes","No"), size = 20, replace = TRUE, prob = c(0.3,0.7)),
                 transport = sample(x = c("shuttle bus", "bicycle", "car"), size = 20, replace = TRUE, prob = c(0.55,0.2,0.25)),
                 math = round(rnorm(n = 20, mean = 7.2, sd = 0.6), 1),
                 english = round(rnorm(n = 20, mean = 8.1, sd = 0.5), 1)
)
```

## Categorical data: frequencies/crosstabs

```{r}
#two-way table
df1 <- table(df$gender,df$class)
df1
addmargins(df1) #add sum of rows and column
round(prop.table(df1,1)) #proportion by rows and round to 1 digits
round(prop.table(df1,2)) #proportion by columns and round to 1 digits
chisq.test(df1) #H0: no relationship
fisher.test(df1) #H0: no relationship
```

```{r}
#three-way crosstabs
df2 <- xtabs(~gender+class+campus, df)
df2
ftable(df2)
```

```{r}

```

## Descriptive statistics




















